### 2019-10-15 boek.tex
reshuffle
update

% figure packages
\usepackage{tikz}
\usetikzlibrary{calc,shapes,arrows,intersections,shadows}
\tikzset{>=latex}
\usepackage{graphicx}
\usepackage[indention=0.5cm,labelsep=colon,font={sf,small},labelfont={bf,sf}]{caption}
\usepackage[indention=0.5cm,font={sf,small},labelfont={bf,sf}]{subcaption}
\usepackage{float}
\usepackage{multirow}
\usepackage{import}
\usepackage{xcolor}
\usepackage{colortbl}
\usepackage{tcolorbox}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%     CH1: INTRODUCTION     %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Introduction} 
\chaptermark{Introduction} % short title
\label{chap:introduction}
\graphicspath{{./ch1_introduction/}}
\lettergroup{\thechapter}
\begin{quote}
	\textbf{Abstract:} 
	%Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in immunology in a high-throughput and unbiased manner. Starting from a mixture of cells in different stages of a developmental process, unsupervised trajectory inference algorithms aim to automatically reconstruct the underlying developmental path that cells are following. In this review, we break down the strategies used by this novel class of methods, and organize their components into a common framework, highlighting several practical advantages and disadvantages of the individual methods. We also give an overview of new insights these methods have already provided regarding the wiring and gene regulation of cell differentiation. As the trajectory inference field is still in its infancy, we propose several future developments which will ultimately lead to a global and data-driven way of studying immune cell differentiation.
\end{quote}
\vfill
\newpage
\include{ch1_introduction/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%        CH2: DYNGEN        %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{dyngen: benchmarking with \textit{in silico} single cells} 
\chaptermark{dyngen: benchmarking with \textit{in silico} single cells}
\label{chap:dyngen}
\graphicspath{{./ch2_dyngen/}}
\include{ch2_dyngen/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%     CH3: DYNBENCHMARK     %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{dynbenchmark: A comparison of single-cell trajectory inference methods} 
\chaptermark{dynbenchmark: A comparison of single-cell trajectory inference methods.} % short title
\label{chap:dynbenchmark}
\graphicspath{{./ch3_dynbenchmark/}}
\begin{quote}
	\textbf{Abstract:} Trajectory inference approaches analyze genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories. Although more than 70 trajectory inference tools have already been developed, it is challenging to compare their performance because the input they require and output models they produce vary substantially. Here, we benchmark 45 of these methods on 110 real and 229 synthetic datasets for cellular ordering, topology, scalability and usability. Our results highlight the complementarity of existing tools, and that the choice of method should depend mostly on the dataset dimensions and trajectory topology. Based on these results, we develop a set of guidelines to help users select the best method for their dataset. Our freely available data and evaluation pipeline (\href{https://benchmark.dynverse.org}{benchmark.dynverse.org}) will aid in the development of improved tools designed to analyze increasingly large and complex single-cell datasets.
\end{quote}
\vfill
Adapted from:\\
Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y. A comparison of single-cell trajectory inference methods. \textit{Nature Biotechnology} 37, 5 (2019), 547--554. \doi{10.1038/s41587-019-0071-9}.\\
{\footnotesize $^*$ Equal contribution}
\newpage
\include{ch3_dynbenchmark/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%         CH4: DYNO         %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{dyno: A toolkit for inferring and interpreting trajectories} 
\chaptermark{dyno: A toolkit for inferring and interpreting trajectories.} % short title
\label{chap:dyno}
\begin{quote}
	\textbf{Abstract:} Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory. Since 2014, at least 75 tools for trajectory inference have been developed. However, for end-users with a dataset of interest, these methods are difficult to execute and compare, mainly due to high variability in input/output data structures, software requirements, and programming interfaces.

	We developed dyno, a set of software tools that allow a user to easily infer, visualise and interpret single-cell trajectories. First, the user can select the most optimal set of methods based on the size of the dataset, the prior knowledge on the trajectories' topology and other user preferences. Each selected method can then be easily run within a common interface. Because the outputs of these methods are all converted into a common consistent format, the trajectories can be easily interpreted. For this, we provide a complete toolkit: adding a root and labelling the different cell stages, detecting genes which are differentially expressed at different stages of the trajectory, and plotting the trajectory within any dimensionality reduction or heatmap.
\end{quote}
\vfill
Adapted from:\\
\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y. dyno: A toolkit for inferring and interpreting trajectories. \textit{Journal} vol, issue (2019), page--page. doi.\\
{\footnotesize $^*$ Equal contribution}
\newpage
\graphicspath{{./ch4_dyno/}}
\include{ch4_dyno/content}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%       CH5: SCORPIUS       %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{SCORPIUS: Fast, accurate, and robust single-cell pseudotime} 
\chaptermark{SCORPIUS: Fast, accurate, and robust single-cell pseudotime.} % short title
\label{chap:scorpius}
\begin{quote}
	\textbf{Abstract:}
\end{quote}
\vfill
Adapted from:\\
\textbf{Cannoodt, R.}, Saelens, W., Sichien, D., Tavernier, S., Janssens, S. Guilliams, M., Lambrecht, B., De Preter, K., and Saeys, Y. SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development. \textit{Journal} vol, issue (2019), page--page. \doi{10.1101/079509v2}.
\newpage
\graphicspath{{./ch5_scorpius/}}
\include{ch5_scorpius/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%         CH6: BRED         %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{bred: Inferring single cell regulatory networks} 
\chaptermark{bred: Inferring single cell regulatory networks.} % short title
\label{chap:bred}
\begin{quote}
	\textbf{Abstract:}
\end{quote}
\vfill
Adapted from:\\
\textbf{Cannoodt, R.}, Saelens, W., Saeys, Y, and De Preter K. Inferring Single Cell Regulatory Networks. \textit{Journal} vol, issue (2019), page--page. doi.
\newpage
\graphicspath{{./ch6_bred/}}
\include{ch6_bred/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%       CH7: INCGRAPH       %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{incgraph: Optimising regulatory networks} 
\chaptermark{incgraph: Optimising regulatory networks.} % short title
\label{chap:incgraph}
\graphicspath{{./ch7_incgraph/}}
\begin{quote}
	\textbf{Abstract:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
	As part of a topology optimisation process, one could use graphlet counts to iteratively modify a network and keep track of the graphlet counts, in order to achieve certain topological properties.
	Up until now, however, graphlets were not suited as a metric for performing topology optimisation; when millions of minor changes are made to the network structure it becomes computationally intractable to recalculate all the graphlet counts for each of the edge modifications. \\ IncGraph is a method for calculating the differences in graphlet counts with respect to the network in its previous state, which is much more efficient than calculating the graphlet occurrences from scratch at every edge modification made.
	In comparison to static counting approaches, our findings show IncGraph reduces the execution time by several orders of magnitude.
	The usefulness of this approach was demonstrated by developing a graphlet-based metric to optimise gene regulatory networks.
	IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation.
\end{quote}
\vfill
Adapted from:\\
\textbf{Cannoodt, R.}, Ruyssinck, J., Ramon, J., De Preter, K., and Saeys, Y. IncGraph: Incremental graphlet counting for topology optimisation. \textit{PLOS ONE} 13, 4 (2018), e0195997. \doi{10.1371/journal.pone.0195997}.
\newpage
\include{ch7_incgraph/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%   CH9: Self-assessment    %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Self-assessment in trajectory inference} 
\chaptermark{Self-assessment in trajectory inference.} % short title
\label{chap:selfassessment}
\graphicspath{{./ch9_selfassessment/}}
\begin{quote}
	\textbf{Abstract:}
\end{quote}
\vfill
Adapted from:\\
\textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. Self-assessment in trajectory inference. \textit{Journal} vol, issue (2019), page--page. doi.
\newpage
\include{ch9_selfassessment/content}
\clearpage
\section{References}
\printbibliography[heading=none]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%       CH10: Guidelines       %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Essential guidelines for computational method benchmarking} 
\chaptermark{Essential guidelines for computational method benchmarking.} % short title
\label{chap:guidelines}
\graphicspath{{./ch10_guidelines/}}
\begin{quote}
	\textbf{Abstract:} In computational biology and other sciences, researchers are frequently faced with a choice between several computational methods for performing data analyses. Benchmarking studies aim to rigorously compare the performance of different methods using well-characterized benchmark datasets, to determine the strengths of each method or to provide recommendations regarding suitable choices of methods for an analysis. However, benchmarking studies must be carefully designed and implemented to provide accurate, unbiased, and informative results. Here, we summarize key practical guidelines and recommendations for performing high-quality benchmarking analyses, based on our experiences in computational biology.
\end{quote}
\vfill
Adapted from:\\
Weber, L. M., Saelens W., \textbf{Cannoodt, R.}, Soneson, C., Hapfelmeier, A., Gardner, P. P.,
Boulesteix, A., Saeys, Y., and Robinson, M. D. 
Essential guidelines for computational method benchmarking. \textit{Genome Biology} 20, 125 (2019). \doi{10.1186/s13059-019-1738-8}.
\newpage
\include{ch10_guidelines/content}

>>>>

% figure packages
\usepackage{tikz}
\usetikzlibrary{calc,shapes,arrows,intersections,shadows}
\tikzset{>=latex}
\usepackage{graphicx}
\usepackage[indention=0.5cm,labelsep=colon,font={sf,small},labelfont={bf,sf}]{caption}
\usepackage[indention=0.5cm,font={sf,small},labelfont={bf,sf}]{subcaption}
\usepackage{float}
\usepackage{multirow}
\usepackage{import}
\usepackage{xcolor}
\usepackage{colortbl}
\usepackage{tcolorbox}
\usepackage[export]{adjustbox}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%     CH1: INTRODUCTION     %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Introduction} 
\chaptermark{Introduction} % short title
\label{chap:introduction}
\graphicspath{{./ch1_introduction/}}
\lettergroup{\thechapter}
%\begin{quote}
%	\textbf{Abstract:} 
%	Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in a high-throughput and unbiased manner. Starting from a mixture of cells in different stages of a developmental process, unsupervised trajectory inference algorithms aim to automatically reconstruct the underlying developmental path that cells are following. In this review, we break down the strategies used by this novel class of methods, and organize their components into a common framework, highlighting several practical advantages and disadvantages of the individual methods. We also give an overview of new insights these methods have already provided regarding the wiring and gene regulation of cell differentiation. As the trajectory inference field is still in its infancy, we propose several future developments which will ultimately lead to a global and data-driven way of studying immune cell differentiation.
%\end{quote}
\vfill
\newpage
\include{ch1_introduction/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%        CH2: DYNGEN        %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{dyngen: benchmarking with \textit{in silico} single cells} 
\chaptermark{dyngen: benchmarking with \textit{in silico} single cells}
\label{chap:dyngen}
\graphicspath{{./ch2_dyngen/}}
%\begin{quote}
%	\textbf{Abstract:}
%	A common problem of pioneering computational tools is that during their development, there are rarely sufficient datasets available for adequately quantitatively assessing its performance. We developed a simulator of \textit{in silico} single cells aimed at being flexible and extendible enough to be used to help kick-start emerging computational sub-domains.
%	
%\end{quote}
\include{ch2_dyngen/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%     CH3: DYNBENCHMARK     %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{dynbenchmark: A comparison of single-cell trajectory inference methods} 
\chaptermark{dynbenchmark: A comparison of single-cell trajectory inference methods}
\label{chap:dynbenchmark}
\graphicspath{{./ch3_dynbenchmark/}}
%\begin{quote}
%	\textbf{Abstract:} Trajectory inference approaches analyze genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories. Although more than 70 trajectory inference tools have already been developed, it is challenging to compare their performance because the input they require and output models they produce vary substantially. Here, we benchmark 45 of these methods on 110 real and 229 synthetic datasets for cellular ordering, topology, scalability and usability. Our results highlight the complementarity of existing tools, and that the choice of method should depend mostly on the dataset dimensions and trajectory topology. Based on these results, we develop a set of guidelines to help users select the best method for their dataset. Our freely available data and evaluation pipeline (\href{https://benchmark.dynverse.org}{benchmark.dynverse.org}) will aid in the development of improved tools designed to analyze increasingly large and complex single-cell datasets.
%\end{quote}
\vfill
%Adapted from:\\
Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y. A comparison of single-cell trajectory inference methods. \textit{Nature Biotechnology} 37, 5 (2019), 547--554. \doi{10.1038/s41587-019-0071-9}.\\
{\footnotesize $^*$ Equal contribution}
\newpage
\include{ch3_dynbenchmark/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%         CH4: DYNO         %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{dyno: A toolkit for inferring and interpreting trajectories} 
\chaptermark{dyno: A toolkit for inferring and interpreting trajectories.} % short title
\label{chap:dyno}
%\begin{quote}
%	\textbf{Abstract:} Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory. Since 2014, at least 75 tools for trajectory inference have been developed. However, for end-users with a dataset of interest, these methods are difficult to execute and compare, mainly due to high variability in input/output data structures, software requirements, and programming interfaces.
%	
%	We developed dyno, a set of software tools that allow a user to easily infer, visualise and interpret single-cell trajectories. First, the user can select the most optimal set of methods based on the size of the dataset, the prior knowledge on the trajectories' topology and other user preferences. Each selected method can then be easily run within a common interface. Because the outputs of these methods are all converted into a common consistent format, the trajectories can be easily interpreted. For this, we provide a complete toolkit: adding a root and labelling the different cell stages, detecting genes which are differentially expressed at different stages of the trajectory, and plotting the trajectory within any dimensionality reduction or heatmap.
%\end{quote}
%\vfill
%Adapted from:\\
%\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y. dyno: A toolkit for inferring and interpreting trajectories. \textit{Journal} vol, issue (2019), page--page. doi.\\
%{\footnotesize $^*$ Equal contribution}
\newpage
\graphicspath{{./ch4_dyno/}}
\include{ch4_dyno/content}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%       CH5: SCORPIUS       %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{SCORPIUS: Fast, accurate, and robust single-cell pseudotime} 
\chaptermark{SCORPIUS: Fast, accurate, and robust single-cell pseudotime.} % short title
\label{chap:scorpius}
%\begin{quote}
%	\textbf{Abstract:} Recent advances in RNA sequencing enable the generation of genome-wide expression data at the single-cell level, opening up new avenues for transcriptomics and systems biology. A new application of single-cell transcriptomics is the unbiased ordering of cells according to their progression along a dynamic process of interest. SCORPIUS is a toolkit for pseudotemporally ordering single cells along a linear trajectory. Comprehensive benchmarking on linear datasets shows that SCORPIUS consistently obtains higher quality linear trajectories in comparison to state-of-the-art trajectory inference methods. We used SCORPIUS to generate novel hypotheses regarding dendritic cell development, which were subsequently validated in vivo.
%\end{quote}
%\vfill
Adapted from:\\
\textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development. bioRxiv 079509. 2016 Oct.
\newpage
\graphicspath{{./ch5_scorpius/}}
\include{ch5_scorpius/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%         CH6: BRED         %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{bred: Inferring single cell regulatory networks} 
\chaptermark{bred: Inferring single cell regulatory networks.} % short title
\label{chap:bred}
%\begin{quote}
%	\textbf{Abstract:}
%\end{quote}
\newpage
\graphicspath{{./ch6_bred/}}
\include{ch6_bred/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%       CH7: INCGRAPH       %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{incgraph: Optimising regulatory networks} 
\chaptermark{incgraph: Optimising regulatory networks.} % short title
\label{chap:incgraph}
\graphicspath{{./ch7_incgraph/}}
%\begin{quote}
%	\textbf{Abstract:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
%	As part of a topology optimisation process, one could use graphlet counts to iteratively modify a network and keep track of the graphlet counts, in order to achieve certain topological properties.
%	Up until now, however, graphlets were not suited as a metric for performing topology optimisation; when millions of minor changes are made to the network structure it becomes computationally intractable to recalculate all the graphlet counts for each of the edge modifications. \\ IncGraph is a method for calculating the differences in graphlet counts with respect to the network in its previous state, which is much more efficient than calculating the graphlet occurrences from scratch at every edge modification made.
%	In comparison to static counting approaches, our findings show IncGraph reduces the execution time by several orders of magnitude.
%	The usefulness of this approach was demonstrated by developing a graphlet-based metric to optimise gene regulatory networks.
%	IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation.
%\end{quote}
\vfill
%Adapted from:\\
\textbf{Cannoodt, R.}, Ruyssinck, J., Ramon, J., De Preter, K., and Saeys, Y. IncGraph: Incremental graphlet counting for topology optimisation. \textit{PLOS ONE} 13, 4 (2018), e0195997. \doi{10.1371/journal.pone.0195997}.
\newpage
\include{ch7_incgraph/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%   CH9: Self-assessment    %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Self-assessment in trajectory inference} 
\chaptermark{Self-assessment in trajectory inference.} % short title
\label{chap:selfassessment}
\graphicspath{{./ch9_selfassessment/}}
%\begin{quote}
%	\textbf{Abstract:}
%\end{quote}
%\vfill
%Adapted from:\\
%\textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. Self-assessment in trajectory inference. \textit{Journal} vol, issue (2019), page--page. doi.
\newpage
\include{ch9_selfassessment/content}
\clearpage
\section{References}
\printbibliography[heading=none]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%       CH10: Guidelines       %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Essential guidelines for computational method benchmarking} 
\chaptermark{Essential guidelines for computational method benchmarking.} % short title
\label{chap:guidelines}
\graphicspath{{./ch10_guidelines/}}
%\begin{quote}
%	\textbf{Abstract:} In computational biology and other sciences, researchers are frequently faced with a choice between several computational methods for performing data analyses. Benchmarking studies aim to rigorously compare the performance of different methods using well-characterized benchmark datasets, to determine the strengths of each method or to provide recommendations regarding suitable choices of methods for an analysis. However, benchmarking studies must be carefully designed and implemented to provide accurate, unbiased, and informative results. Here, we summarize key practical guidelines and recommendations for performing high-quality benchmarking analyses, based on our experiences in computational biology.
%\end{quote}
\vfill
%Adapted from:\\
Weber, L. M., Saelens W., \textbf{Cannoodt, R.}, Soneson, C., Hapfelmeier, A., Gardner, P. P.,
Boulesteix, A., Saeys, Y., and Robinson, M. D. 
Essential guidelines for computational method benchmarking. \textit{Genome Biology} 20, 125 (2019). \doi{10.1186/s13059-019-1738-8}.
\newpage
\include{ch10_guidelines/content}

### 2019-10-15 ch0_frontmatter.tex
update
~
reshuffle

%%%%%%%%%%%%%%%%%%%%%%%%
%%% DEFINE TITLEPAGE %%%
%%%%%%%%%%%%%%%%%%%%%%%%
\begin{titlepage}
	\centering
	\ 
	\vfill
	{\Huge \textsf{Modelling single cell dynamics\\with trajectories and gene regulatory networks}}

@@

\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Acknowledgements}
\blindtext

@@

\newcommand{\grn}{GRN}
\newcommand{\grns}{GRNs}
\nomenclature{GRN}{Gene Regulatory Network}

\newcommand{\im}{IM}
\newcommand{\ims}{IMs}
\nomenclature{IM}{Importance Measure}

\newcommand{\tf}{TF}
\newcommand{\tfs}{TFs}
\nomenclature{TF}{Transcription Factor}

\newcommand{\dna}{DNA}
\nomenclature{DNA}{Deoxyribonucleic Acid}

\newcommand{\rna}{RNA}
\nomenclature{RNA}{Ribonucleic Acid}

\newcommand{\mrna}{mRNA}
\nomenclature{mRNA}{Messenger RNA}

\newcommand{\nin}{NI}
\nomenclature{NI}{Network Inference}

\newcommand{\ml}{ML}
\nomenclature{ML}{Machine Learning}

\newcommand{\rf}{RF}
\nomenclature{RF}{Random Forests}

\newcommand{\cart}{CART}
\nomenclature{CART}{Classification And Regression Trees}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%
%%% DEFINE TITLEPAGE %%%
%%%%%%%%%%%%%%%%%%%%%%%%
\begin{titlepage}

	\includegraphics[height=1.5cm]{fig/logos/logo_UGent_EN_RGB_color}

	\centering

	\vfill
	{\Huge \textsf{Modelling single cell dynamics\\with trajectories and gene regulatory networks}}

@@

%\newpage{\thispagestyle{empty}\cleardoublepage}
%\chapter*{Acknowledgements}

@@

\nomenclature{HCA}{Human Cell Atlas}
\nomenclature{TCGA}{The Cancer Genome Atlas}

\nomenclature{DNA}{Deoxyribonucleic Acid}
\nomenclature{RNA}{Ribonucleic Acid}
\nomenclature{mRNA}{Messenger RNA}
\nomenclature{TF}{Transcription Factor}

\nomenclature{GRN}{Gene Regulatory Network}
\nomenclature{NI}{Network Inference}
\nomenclature{CART}{Classification And Regression Trees}
\nomenclature{RF}{Random Forests}
\nomenclature{TI}{Trajectory Inference}

\nomenclature{$k$NN}{$k$-nearest-neighbour}
\nomenclature{DR}{Dimensionality Reduction}
\nomenclature{PCA}{Principal Component Analysis}
\nomenclature{MDS}{Multi-Dimensional Scaling}
\nomenclature{t-SNE}{t-distributed Stochastic Neighbor Embedding}
\nomenclature{UMAP}{Uniform Manifold Approximation and Projection}
\nomenclature{LDMS}{Landmark Multi-Dimensional Scaling}
\nomenclature{DE}{Differential Expression}
\nomenclature{TDE}{Trajectory Differential Expression}

\chapter*{Summary}

Recent technological developments triggered a major milestone in cellular biology by being able to profile and monitor biomolecular contents of single cells at a genome-wide level. The rapid advancements in single-cell omics allow for new types of analyses that fundamentally redefine our understanding of cellular heterogeneity and dynamic processes. Even now, there are even more exciting advancements on the horizon, with the Human Cell Atlas aiming to create a biomolecular and spatial map of all the different cell types in the body.

Throughout this doctoral project, I did not only get to follow these inspiring advancements first hand, but was able to make scientific contributions to the field by extensively reviewing novel computational tools, and developing several tools and algorithms of my own. More specifically, my efforts have resulted in the following contributions.

A common problem of pioneering computational tools is that during their development, there are rarely sufficient datasets available for adequately quantitatively assessing its performance. We developed a simulator of \textit{in silico} single cells aimed at being flexible and extendible enough to be used to help kick-start emerging computational sub-domains (Chapter~\ref{chap:dyngen}).

We performed a comprehensive comparison of 45 trajectory inference methods (Chapter~\ref{chap:dynbenchmark}). Trajectory inference is a new type of computational analysis whose aim is to identify and study transitions between main cellular states. While trajectory inference is one of the most rapidly growing new fields in single-cell omics tools, a quantitative study of the advantages and disadvantages was hitherto lacking.

We developed four new software tools ourselves. These methods include an extensive toolkit for inferring, visualising and interpreting trajectories (Chapter~\ref{chap:dyno}), a robust tool for inferring linear trajectories (Chapter~\ref{chap:scorpius}), a novel way for studying gene regulation at a single-cell or single-sample level (Chapter~\ref{chap:bred}), and tool for quantifying the topological changes resulting from minor perturbations in a network (Chapter~\ref{chap:incgraph}).

We reflect on current challenges within the field, by providing guidelines for how to self-assess trajectory inference methods (Chapter~\ref{chap:selfassessment}) and, more generally, how to perform large-scale benchmarking experiments of computational methods (Chapter~\ref{chap:guidelines}).

Throughout this project I particularly enjoyed working in an open-source ecosystem, collaborating with scientists around the world to tackle unsolved problems, whilst learning best practices by inspecting the source code of exemplary software projects.

\chapter*{Samenvatting}

Recente technologische ontwikkelingen hebben geleid tot significante ontwikkelingen in celbiologie die ons toelieten om de biomoleculaire inhoud van individuele cellen op een genoomwijde manier te observeren. Dankzij snelle ontwikkelingen in \textit{single-cell omics} technologieën wordt ons verstaan van cellulaire heterogeniteit en dynamische processen voortdurend gewijzigd. Inmiddels zijn er nieuwe spannende ontwikkelingen onderweg, zoals het \textit{Human Cell Atlas} project, die mikt om alle celtypes in het menselijke lichaam op biomoleculaire en spatiële wijze in kaart te brengen.

Gedurende dit doctoraatsproject heb ik deze inspirerende ontwikkelingen niet alleen van heel dichtbij kunnen volgen, maar ik heb ook wetenschappelijke bijdragen aan het veld kunnen leveren door nieuwe analytische programma’s extensief te onderzoeken, en door zelf verscheidene algoritmes te ontwikkelen. Meer specifiek heeft mijn onderzoek geleid tot de volgende bijdragen.

Een recurrent probleem van baanbrekende computationele algoritmes is dat gedurende hun ontwikkeling er helaas vaak onvoldoende datasets beschikbaar zijn om de performantie op een kwantitatieve wijze te kunnen beoordelen. We ontwikkelden een programma om individuele cellen te simuleren, dat flexibel en uitbreidbaar genoeg is om toekomstige computationele onderzoeksvelden te ondersteunen (Hoofdstuk~\ref{chap:dyngen}).

We voerden een comprehensieve studie van 45 \textit{trajectvoorspellende} methoden uit (Hoofdstuk~\ref{chap:dynbenchmark}). Trajectvoorspellende methoden is een nieuwe soort van computationele analyse waarbij het doel is om transities tussen verscheidene celtypes te identificeren en bestuderen. Terwijl deze klasse van methoden een van de snelst groeiende subdomeinen binnen \textit{single-cell omics} was, was er tot op dat moment nog geen kwantitatieve studie van de voor- en nadelen van dergelijke methoden uitgevoerd.

We ontwikkelde ook zelf vier nieuwe computer programma's, waaronder een uitgebreide toolkit om trajecten te voorspellen, visualiseren en interpreteren (Hoofdstuk~\ref{chap:dyno}), een robuust programma om lineaire trajecten te voorspellen (Hoofdstuk~\ref{chap:scorpius}), een nieuwe manier om gen regulatie op het niveau van een individuele cel of staal te bestuderen (Hoofdstuk~\ref{chap:bred}), en een programma om topologische veranderingen te kwantificeren ten gevolge van kleine veranderingen in een netwerk (Hoofdstuk~\ref{chap:incgraph}).

We reflecteren op huidige uitdagingen binnen het veld, door richtlijnen aan te bieden over hoe je je eigen trajectvoorspellende methode kan evalueren (Hoofdstuk~\ref{chap:selfassessment}) en, in het algemeen, hoe je grootschalige beoordelingen van computationele methoden kan uitvoeren (Hoofdstuk~\ref{chap:guidelines}).

Tijdens het verloop van dit project heb ik bijzonder genoten van te werken in een \textit{open-source} omgeving, waarbij ik wereldwijd collaboraties kon aangaan met onderzoekers. Hierbij heb ik bijgeleerd door naar de broncode van voorbeeldige softwareprojecten te kunnen kijken.
